---
title: "Reglas de Asociación - Implicaciones"
description: |
  Realization of Proyecto III - reglas, FCA.
author:
  #"Jakub Maciążek"
  name: "Jakub Maciążek </br>"
  affiliation: Faculty of Information and Communication Technology, Wrocław University of Science and Technology
subtitle: "Proyecto III"
output:
  rmdformats::downcute:
    use_bookdown: true
    code_folding: show
    self_contained: true
  pdf_document:   
    latex_engine: xelatex   # added to knit to pdf 
  html_document: default    # added to knit to html 
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(ggplot2)
library(latex2exp)
knitr::opts_chunk$set(fig.align = "center",
                      fig.width = 5,
                      fig.height = 4,
                      collapse = TRUE)



library(readr)
library(arules)
library(arulesViz)
library(fcaR)
library(hasseDiagram)
```

# Project description

**Social Network Analysis**

  For this project we will use the data obtained from Twitter used for a research article that we published in the Knowledge-Based Systems Journal on the topic of social network analysis using FCA. The objective was to identify leaders of the social network.
  
  - download dataset
  - dataset stores topics (rows) treated by means of communication (columns)
  - process in what you consider (reduce, discretize, group, etc.) - justify if you do it due to excessive computing time, etc.
  
**Basic analysis**

  - Extract patterns or association rules using arules
  - Extract knowledge using fcaR
  
  It's not about copy-paste commands. Data analysis involves extracting useful knowledge, drawing conclusions. Extract patterns, visualize, show hidden knowledge with any technique from the themes of implications and association rules.

Explain the steps you are taking.

**Going deeper**

  - Open section for you to think of something interesting that you are able to extract from the dataset analysis using rules or FCA.
  - Identify media that really act as leaders? either
  - What are the most important topics? either
  - Is there interesting knowledge in the lattice? On the left side of the implications? attributes that are repeated a lot in association rules?, …
 
 
# Basic analysis

## Initial data analysis

  We will begin with data import.
  
```{r}
sn <- read.csv("S:/0_Universidad_de_Malaga/MI_Ingenieria_y_ciencia_de_datos/Estatistica_avanzada_para_ciencia_de_datos/Reglas_de_asociacion/proyecto_tercera_parte/5000 de 26 medios completo-1.csv",  sep=";")
head(sn)

rows_labels <- sn[ , c("X")]
head(rows_labels)
sn <- subset(sn, select = -c(X) )
head(sn)
rownames(sn) <- rows_labels
head(sn)
```

  Removal of columns where all values are 0 (Media didn't occure in any topic).
  
```{r}
#sn_processed <- sn[, which(colSums(sn) != FALSE)]
```

  Given data is already discreet, therefore does not require any transformations, except generation of set of transactions for *arules* analysis.
  
```{r}
#trans1 <- transactions(sn)
#trans1
#inspect(trans1)
```
  
## Arules analysis

## Formal Concept Analysis

```{r}
fc_sn <- FormalContext$new(sn)
fc_sn$print()
```

```{r}
fc_sn$dim()

fc_sn$attributes

fc_sn$objects

fc_sn$plot()
```
```{r}
fc_sn$find_concepts()
fc_sn$concepts$size()
head(fc_sn$concepts)
```

```{r}
# Wrong version of R

#fc_sn$concepts$plot(object_names = FALSE)
```

```{r}
fc_sn$concepts$support()

idx <- which(fc_sn$concepts$support() > 0.2)

sublattice <- fc_sn$concepts$sublattice(idx)
sublattice
```

# Deeper analysis




