---
title: "Series Temporales"
description: |
  Realization of time series exercise.
author:
  #"Jakub Maciążek"
  name: "Jakub Maciążek </br>"
  affiliation: Faculty of Information and Communication Technology, Wrocław University of Science and Technology
subtitle: "Mini-tutorial on the use of zoo and xts packages"
output:
  rmdformats::downcute:
    use_bookdown: true
    code_folding: show
    self_contained: true
  html_document: default    # added to knit to html 
  pdf_document:   
    latex_engine: xelatex   # added to knit to pdf 
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(ggplot2)
library(latex2exp)
knitr::opts_chunk$set(fig.align = "center",
                      fig.width = 5,
                      fig.height = 4,
                      collapse = TRUE)


library(TTR)
library(forecast)
library(readr)
library (dplyr)
```

# Exercise description

  Make an .rmd with a mini-tutorial on the use of the zoo and xts packages, applying them to a time series (dataset with data from Spain) that you search on the web about electricity consumption, product consumption, etc.

# Data source

  Following tutorial will be based on data related to electricy consumption, specifically daily energy SPOT price, from the years 2014 - 2018.
  
  Dataset source: https://www.kaggle.com/datasets/manualrg/spanish-electricity-market-demand-gen-price.

```{r}
#Load data

spain_energy_market <- read.csv("S:/0_Universidad_de_Malaga/MI_Ingenieria_y_ciencia_de_datos/Estatistica_avanzada_para_ciencia_de_datos/Series_temporales/ejercicio/spain_energy_market.csv")
spain_SPOT_prices <- filter(spain_energy_market, name=='Precio mercado SPOT Diario ESP')
head(spain_SPOT_prices)

keeps <- c("datetime", "value")
data <- spain_SPOT_prices[keeps]
head(data)
```
  
# Brief package introduction

**xts** - eXtensible Time Series

**zoo** object - index + matrix -> observation in time

# Functionalities

  First basic functionalities are presented. Those include creation of xts objects, and conversion back to raw data frame.

## Basics

```{r}
#Import package
library(xts)

# Construct xts
x <- xts(data$value, order.by = as.Date(data$datetime))
head(x)

# Get back raw matrix
head(coredata(x, fmt = FALSE))

# Extract index/time
head(index(x))

# Class attributes - index class
indexClass(x)
```

  Following, multiple functions for solving the problem of missing values are presented.

```{r}
#Omit NA values in x 
x_ommited <- na.omit(x)

#Fill missing values in x using last observation
x_na_last <-na.locf(x)

#Fill missing values in x using next observation
x_na_next <- na.locf(x,  fromLast=TRUE)

#Interpolate NAs
x_na_interpolated <- na.approx(x) 
```

  Lastly, some general functions are listed.

```{r}
#Value of data_xts in index of data_xts
head(.indexwday(x))

#First and last observation
start(x)
end(x)

# Data structure of x
str(x)
```

## Data import / export

  XTS objects can not only be created, but also imported and the converted. Following are listed multiple ways to import and export data, including efficient usage of zoo.

```{r}
# Convert imported data to xts (needs to contain datetime class variable)
data$datetime <- as.Date(data$datetime)
data_xts <- as.xts(data)
head(data_xts)

# class type
class(data_xts)
```

```{r}
#Read data using zoo
#as.xts(read.zoo("file_name"))

# Save to external file
write.zoo(data_xts, file = "xts_file_name.csv", sep = "'")

# Save for R use - optimized for objects like xts -> fast read
saveRDS(data_xts, "xts_RSD_file_name.rds")
head(readRDS("xts_RSD_file_name.rds"))
```

## Periods and periodicity

  Time series analysis is inherently related to the time space/intervals in which data was observed. Following listed are functions related to periodic nature of TS.

Estimation of frequency and time-span of observations.
```{r}
periodicity(data_xts)
```

Conversion of xts to monthly, yearly OHLC. 

OHLC - first (Opening) and last (Close) value from period, plus Highest and Lowest
```{r}
data_xts_monthly <- to.monthly(data_xts)
head(data_xts_monthly)
#Count months
nmonths(data_xts_monthly)

data_xts_yearly <- to.yearly(data_xts)
data_xts_yearly
```

Conversion of xts to monthly periods, while aggregating values by mean.
```{r}
data_xts_monthly_mean <- apply.monthly(data_xts, mean)
tail(data_xts_monthly_mean)
```

Duplicates removal
```{r}
data_xts_unique <- make.index.unique(data_xts,drop=TRUE)
head(data_xts_unique)
```


## Visualization

  Finally, this section presents visualization of xts. By nature, xts objects can skip xts in plot.xts, which is required for standard data frames.

```{r}
# If plotted object is xts object, simple plot() can be used
plot.xts(data_xts, main = "Main title")
plot(data_xts_monthly_mean)
lines(data_xts_monthly_mean, col = "green", lwd = 2) #lwd - line fitnes adjustment
```


<!--
Course:

  - https://www.datacamp.com/cheat-sheet/xts-cheat-sheet-time-series-in-r
  - R tutorial: xts & zoo for time series analysis: https://www.youtube.com/watch?v=qXShXRdYofo
  
-->
